---
title: "IE 5374 - Section 1 - Group 5 - Project-1"
author: "Naren Mohan, Simran Bhatia, Srujana Gaddam"
date: "11/21/2021"
output: html_document
---

# Objective
The objective of project 1 is to implement different clustering methods to synthetic and real-world data and validate using external and internal validation techniques

# Importing all required libraries
```{r}
library(readr)
library(ClusterR)
library(cluster)
library(ggplot2)
library(plotly)
library(factoextra)
```

# Importing all the data
```{r}
data1 <- read_csv("IE5374/Homeworks/Project 1/Data1.csv", show_col_types = FALSE)
data2 <- read_csv("IE5374/Homeworks/Project 1/Data2.csv", show_col_types = FALSE)
data3 <- read_csv("IE5374/Homeworks/Project 1/Data3.csv", show_col_types = FALSE)
data4 <- read_csv("IE5374/Homeworks/Project 1/Data4.csv", show_col_types = FALSE)
data5 <- read_csv("IE5374/Homeworks/Project 1/Data5.csv", show_col_types = FALSE)
data6 <- read_csv("IE5374/Homeworks/Project 1/Data6.csv", show_col_types = FALSE)
data7 <- read_csv("IE5374/Homeworks/Project 1/Data7.csv", show_col_types = FALSE)
data8 <- read_csv("IE5374/Homeworks/Project 1/Data8.csv", show_col_types = FALSE)
world_indicators <- read_csv("IE5374/Homeworks/Project 1/World Indicators.csv", show_col_types = FALSE)
```

# Pre-processing some datasets
```{r}
names(data2) <- c("...1", "X1", "X2", "X3", "Class")
data2$...1 <- gsub("X", "", as.character(data2$...1))
data2$...1 <- as.numeric(data2$...1)
data2 <- data2 %>%
  na.omit()

data6$X3 <- rep(1, times=dim(data6)[1])
data7$X3 <- rep(1, times=dim(data7)[1])
```

# Setting a seed value
```{r}
set.seed(4)
```

# Task 1
Data sets “Data1.csv”, “Data2.csv”, “Data3.csv”, “Data4.csv”, “Data5.csv”, “Data6.csv”, “Data7.csv”, “Data8.csv” contain the data points and their respective class information.

For each of the datasets follow the below steps:
1. Use K-means and hierarchical clustering methods to generate clusters
2. Evaluate the performance of the clustering algorithm using external validation metrics
3. Plot (2D or 3D) the data points for each dataset and color them according to the original class
4. Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

## Creating a function to perform k-means on input dataset using input no of clusters
```{r}
performKClustering <- function(dataset, nClusters) {
  kMeans <- kmeans(dataset, centers=7, 10000, nstart=10)
  kClusters <- kMeans$cluster
  kCenters <- as.data.frame(kMeans$centers)

  plot <- plot_ly(data = dataset,
                  x=dataset$X1,
                  y=dataset$X2,
                  z=dataset$X3,
                  type="scatter3d",
                  mode="markers",
                  color=as.factor(kClusters),
                  colors="Set1",
                  alpha=0.7,
                  showlegend=FALSE)
  
  plot <- add_trace(plot, 
                x=kCenters$X1,
                y=kCenters$X2,
                z=kCenters$X3, 
                type="scatter3d",
                mode="markers",
                color=as.factor(rep(1, 7)),
                colors=c("#000000"),
                stroke=I("Black"),
                span=I(1))

  plot <- plot %>%
    layout(title = "K means clustering of Dataset",
           scene = list(
             xaxis = list(title = 'X1'),
             yaxis = list(title = 'X2'),
             zaxis = list(title = 'X3')))
  
  plot
  
  return(kClusters)
}
```

## Creating a function to perform hierarchical clustering on input dataset using input no of clusters
```{r}
performHClustering <- function(dataset, nClusters) {
  dist <- dist(dataset, method = "euclidean")
  hc <- hclust(dist, method = "complete" )
  plot(hc, cex = 0.6, hang = -1)
  hclust <- cutree(hc, k = nClusters)
  
  plot <- plot_ly(data = dataset,
                  x=dataset$X1,
                  y=dataset$X2,
                  z=dataset$X3,
                  type="scatter3d",
                  mode="markers",
                  color=as.factor(hclust),
                  colors="Set1",
                  alpha=0.7,
                  showlegend=FALSE)
  
  plot <- plot %>%
    layout(title = "Hierarchical clustering of Dataset",
           scene = list(xaxis = list(title = 'X1'),
                        yaxis = list(title = 'X2'),
                        zaxis = list(title = 'X3')))
  
  plot
  
  return(hclust)
}
```

## Creating a function to analyze data and review original clusters
```{r}
analyzeDataset <- function(dataset) {
  plot <- plot_ly(data = dataset,
                 x=dataset$X1,
                 y=dataset$X2,
                 z=dataset$X3,
                 type="scatter3d",
                 mode="markers",
                 color=as.factor(dataset$Class),
                 alpha = 0.7)
  plot <- plot %>%
    layout(title = "Dataset with True Labels",
           scene = list(xaxis = list(title = 'X1'),
                        yaxis = list(title = 'X2'),
                        zaxis = list(title = 'X3')))
  plot
  
  # removing first and last columns from input dataset and adding them to a new variable
  subdataset <- dataset %>%
    select(-c("...1", "Class"))

  return(subdataset)
}
```

## Dataset 1 - Data1.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data1$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data1, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of 5 and 8.

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data1, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of 8.

```{r}
data1_1 <- analyzeDataset(data1)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data1_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data1$Class, kClusters, summary_stats = TRUE)
```
We are looking at nearly ideal values above which shows that the performance of the clustering algorithm has been decent

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data1_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data1$Class, hClusters , summary_stats = TRUE)
```
We are looking at ideal values above because the clustering algorithm has perfectly clustered the given dataset.

## Dataset 2 - Data2.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data2$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data2, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of 4.

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data2, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of 2.

```{r}
data2_1 <- analyzeDataset(data2)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data2_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data2$Class, kClusters, summary_stats = TRUE)
```
We are looking at nearly ideal values above which shows that the performance of the clustering algorithm has been decent

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data2_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data2$Class, hClusters , summary_stats = TRUE)
```
Here we can observe that the Hierarchical Clustering method is performing poorly in comparison with k-means clustering method

## Dataset 3 - Data3.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data3$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data3, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of 4.

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data3, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of 2.

```{r}
data3_1 <- analyzeDataset(data3)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data3_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data3$Class, kClusters, summary_stats = TRUE)
```
We are looking at ideal values above which shows that the performance of the clustering algorithm has been the best

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data3_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data3$Class, hClusters , summary_stats = TRUE)
```
Here we can observe that the Hierarchical clustering method is performing poorly in comparison with k-means clustering method

## Dataset 4 - Data4.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data4$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data4, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of ENTER TEXT HERE

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data4, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of ENTER TEXT HERE

```{r}
data4_1 <- analyzeDataset(data4)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data4_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data4$Class, kClusters, summary_stats = TRUE)
```
We are looking at ideal values above which shows that the performance of the clustering algorithm has been the best

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data4_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data4$Class, hClusters , summary_stats = TRUE)
```
The dataset is quite complex for both K means and Hierarchical (since both are distance based) algorithms. We can see that the performance of the algorithm is poor

## Dataset 5 - Data5.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data5$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data5, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of ENTER TEXT HERE

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data5, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of ENTER TEXT HERE

```{r}
data5_1 <- analyzeDataset(data5)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data5_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data5$Class, kClusters, summary_stats = TRUE)
```
ENTER TEXT HERE

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data5_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data5$Class, hClusters , summary_stats = TRUE)
```
ENTER TEXT HERE

## Dataset 6 - Data6.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data6$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data6, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of ENTER TEXT HERE

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data6, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of ENTER TEXT HERE

```{r}
data6_1 <- analyzeDataset(data6)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data6_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data6$Class, kClusters, summary_stats = TRUE)
```
ENTER TEXT HERE

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data6_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data6$Class, hClusters , summary_stats = TRUE)
```
ENTER TEXT HERE

## Dataset 7 - Data7.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data7$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data7, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of ENTER TEXT HERE

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data7, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of ENTER TEXT HERE

```{r}
data7_1 <- analyzeDataset(data7)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data7_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data7$Class, kClusters, summary_stats = TRUE)
```
ENTER TEXT HERE

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data7_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data7$Class, hClusters , summary_stats = TRUE)
```
ENTER TEXT HERE

## Dataset 8 - Data8.csv

### Internal validation and analysis of the cluster
How many unique classes exist in this data set?
```{r}
nClusters <- length(unique(data8$Class))
cat("Number of classes in the dataset:", nClusters)
```

What is the ideal number of clusters using WSS method?
```{r}
fviz_nbclust(data8, FUNcluster = kmeans, method = "wss")
```
WSS method suggests an optimal number of clusters of ENTER TEXT HERE

What is the ideal number of clusters suggested by silhouette method?
```{r}
fviz_nbclust(data8, FUNcluster = kmeans, method = "silhouette")
```
Average silhouette method suggests an optimal number of clusters of ENTER TEXT HERE

```{r}
data8_1 <- analyzeDataset(data8)
```

### Using k-means clustering
```{r}
kClusters <- performKClustering(data8_1, nClusters)
```

### External validation of k-means
```{r}
external_validation(data8$Class, kClusters, summary_stats = TRUE)
```
ENTER TEXT HERE

### Hierarchical Clustering
```{r echo=TRUE, eval=TRUE}
hClusters <- performHClustering(data8_1, nClusters)
```

### External validation of hierarchical clustering
```{r echo=TRUE, eval=TRUE}
external_validation(data8$Class, hClusters , summary_stats = TRUE)
```
ENTER TEXT HERE